Analysis of protein tunnel traversability for flexible ligands

### 2017-05-04 paper.tex
added algorithm

Ligands are small molecules interacting with the protein. 
Ligands are not rigid structures, they rather change due to interactions with the protein's atoms.
Flexible ligands can be modeled as a kinematic chain with predefied degrees of freedom.
An alternative approach, that is used in this work, is to consider a predefined set of possible conformations.
The conformations are prepared using Rosetta software TODO. and their are denoted $c$.

%We further consider a spherical probe $\Sprobe$  of radius $\probe$ encapsulating the ligand.
Let $q=(x,y,z)\in\C$ denote the configuration (position) of the probe, where
the configuration space $\C$ consists of all possible configurations.
For each configuration $q\in\C$ we assume, that its largest collision-free radius $r(q) \in \mathbf{R}, r(q)\ge 0$ can be computed, e.g.
using collision detection.
%using collision detection.
%Let $L(q) \subseteq \mathbf{R}^3$ denote the geometry of the ligand at position $q \in \C$.
%Similarly to geometry of molecule, the geometry of ligand is union of hard sphere model.
%A probe is represented by one sphere of radius $\probe$.
The collision-free region $\CF \subseteq \C$ is formed by configurations, where $\Sprobe$ can be placed without any collision, i.e., 
$\CF = \{q \in \C | \Sprobe(q) \cap \SS = \emptyset\}$, where $\Sprobe(q)$ is the sphere with radius $\probe$ at configuration $q$.
%We assume, that atoms located on the surface of the molecule are known, these are reffered to as {\sl surface atoms}.
The distance $\dist(q_1,q_2)$  between two configurations $q_1,q_2\in\C$ is measured using 3D Euclidean metric.

%This radius can be detected using collision detection method.
The configurations outside the protein $\CG \subseteq \C$ are collision-free with sphere $\Sgprobe$ of radius $\gprobe$, $\gprobe> \probe$,
$\CG=\{q\in \C| \Sgprobe(q) \cap \SS = \emptyset \}$.
%A configuration $q \in \CG$ is referred to as {\sl goal configuration} in the rest of the paper.
By computing $\alpha$-shape of the molecule, we can identify atoms located at the surface. 
These atoms are referred to as {\sl surface atoms} in the rest of the paper.
The distance between a configuration $q \in \C$ and its nearest surface atom is referred to as $\dists(q)$.
%We assume, that for each configuration $q$, the largest collision-free radius $r(q)$ can be computed.
%Let $r(q)$ denote the radius of the largest collision-free sphere at configuration $q$.

Dynamic proteins are represented by a sequence of $n$ snapshots (frames).
The tunnels change with the protein dynamics as well: their width and position change, they can merge with other tunnels
or even disappear.
The free region $\CF$ is therefore different in each frame, as positions of atoms change.
For the sake of simplicity, a single notation $\C$ resp. $\CF$ is used in this paper and it is valid for the frame being processed.

Approximated WVD diagram of the atoms is computed by representing each atom using 12 balls of equal radii and computing ordinary Voronoi diagram.
Computing WVD this way is more efficient and numerically stable and it is used also in other related tools~\cite{caver3,yaffe2008}.
Vertices of this WVD are filtered out if their distance to an atom is less than probe radius $\probe$.
Let $\VV$ contain the remaining Voronoi vertices, and let $r_v \in \VV$ denote the distance to the nearest atom.
%The remaining Voronoi vertices $\VV$ of the WVD and their distances $r_v, v\in \VV$ to the nearest atom are then used in the following method.

%\section{Tunnel detection}

%The tunnel detection using motion planning leads to search in 4-dimensional $\C\!\times\!time$ space, which can be searched using randomized sampling-based methods.
%The task of tunnel detection is to find a collision-free path from a given starting point located inside a dynamic protein 
%to the outer environment. %: for a probe $\Sprobe$.
%The idea of the proposed approach is to detect tunnels in each frame separately using a novel modification of RRT algorithm.
%The configuration tree built in a given frame is transferred to the next frame, where the search continues.

%We propose to detect the tunnels in this 4D space
%(the space consists of pairs $(q,t) \in \C \times [1,n]$, where $n$ is the number of frames) by randomized sampling.
%where $\C$ is three-dimensional configuration space
%of the spherical probe and $n$ is the number of frames.
%It is not computationally feasible to discretize the space to a grid, in which a path is found using standard space-search methods.
%Moreover, finding a collision-free path in the space does not solve the problem, because the probe cannot move back in time.
%Tunnel detection in dynamic proteins leads to search in $\C \times time$ space. %~\cite{Kindel00kinodynamic}.
%sometimes called state-time space~\cite{berg2005roadmap}.
%The proposed solution samples the configuration space of each frame separately in order to detect possible tunnels in the frame.
%The resulting tree is then pruned to remove collision with atoms of the next frame.

@@

\linesnumbered
\begin{algorithm}[h]
%\setstretch{\straa}
{\small
\caption{\label{alg::tunnel}static tunnel detection}
\KwIn{Initial configuration tree $\T$,
    initial set of blocking spheres $\SB'$
}
\KwData{goal-probe radius $\gprobe$,
    probe radius $\probe$, 
    max. number of iterations $\Imax$, 
    phase switching threshold $\alpha$,
    threshold $\ths$,
    num. of new nodes $m$ in each iteration,
    prob. $\gb$ of sampling randomly from $\C$
}
\KwOut{Configuration tree $\T$,
    list of tunnel exit-points $G$
}
\hrule
$\SB = \SB'$; //blocking atoms added during phase~1 \\
$G = \emptyset$; // list of tunnel exit-points \\
$phase=1$\;
\For{$iteration = 1:\Imax$}{
    \If(// Voronoi-based sample --- Section~\ref{sec::vbg}){$phase=1$}{
        $\VVA =$ update active Voronoi points from $\VV$\;
        $v =$ random point from $\VVA$\;
        $\qrand =$ random point around $v$ in the distance $r_v$\;
%        $a_1$ = random atom on surface\;
%        $a_2$ = random atom on surface\;
%        $\qrand$ = random point on line $|a_1, a_2|$\;
    }
    \If{$rand() < \gb$}{
        $\qrand$ = random sample from $\C$\;
        $v = \emptyset$; //no Voronoi vertex used
    }

    $\qnear$ = nearest node in $\T$ towards $\qrand$\;

    \If{$v \ne \emptyset$ {\bf and} $\dist(\qnear, v) < \probe$} {
        add $v$ to $\T$, set its parent to $\qnear$\;
    }

    $N\!=\!m$ collision-free nodes around $\!\qnear$ in the distance $\!\varepsilon$\;
%    $(\qnew,r(\qnew))$ = expand($\T$,$\qnear$, $\SS \cup \SB$)\;
    \For{$q \in N $}{
        \eIf{$\dists(q) < \ths $ {\bf and} $phase=1$}{
            $\SB = \SB \cup \{\Sgprobe(q)\}$ \;
        }{
            $q'$ = nearest node in $\T$ towards $q$\;
            \If{$\dist(q,q') > \dist(q,\qnear)$}{
                add $q$ to $\T$, set its parent to $\qnear$\;
            }
        }
        \If{ $r(q) > \gprobe$ }{
            $G = G \cup \{ q \} $\;
        }
    }
    \If{$phase = 1$ {\bf and} $iteration > \alpha \Imax$}{
        $phase = 2$\;
        $\SB = \emptyset$; // block. atoms are not used in phase 2 \\
    }
}
\return $(\T, G)$\; 
}
\end{algorithm}

>>>>

Ligands are small molecules interacting with the protein. 
Due to the interaction, the relative positions of the ligand's atoms can change.
the ligands  can be modeled as a multi-link kinematic chain where the torional angles phi/psi can change in a given range~\cite{songPFpath}.
After the torsional angles are set, 3D position of each atom of the ligand is computed.
By changing the angles, various conformations can be generated.
However, not all conformations are feasible, which has to be checked using potential energy, which is time consuming.

To decrease the computational burgen, a set of feasible conformations $C$ is prepared.
The conformations can be prepared using Molecular Dynamics software such as Rosetta.
For each conformation $c\in C$ an energy potential is also calculated.
The transition matrix $M$ defines the possible changes of conformations, i.e., $M(i,j)=1$ if the conformation $c_i \in C$
can be switched to the confomration $c_j \in C$.

The proteins are dense structures with narrow tunnels.
Depending on the protein, the tunnel bottlenecks can be even less then $1$~\AA, which is too narrow for ligands with more than 2 atoms.
For the purpose of traversability study, the radii of the ligands are scaled down by a factor $s$.

A configuration of the pseudoflexible ligand $q=(x,y,z,r_x,r_y,r_z,c,s)$ is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (usually it's geometric center), rotation around $x$, $y$ and $z$ axes,
index of the conformation $c$ and scale of atom radii $s$.
All possible configurations form the configuration space $\C$. 
Let $\CF \in \C$ denote the set of collision-free configurations. 
A configuration is collision-free if there is no intersection between the ligand's atoms scaled by the actual scale factor and the protein atoms.

%We further consider a spherical probe $\Sprobe$  of radius $\probe$ encapsulating the ligand.
%Let $q=(x,y,z)\in\C$ denote the configuration (position) of the probe, where
%the configuration space $\C$ consists of all possible configurations.
%For each configuration $q\in\C$ we assume, that its largest collision-free radius $r(q) \in \mathbf{R}, r(q)\ge 0$ can be computed, e.g.
%using collision detection.
%using collision detection.
%Let $L(q) \subseteq \mathbf{R}^3$ denote the geometry of the ligand at position $q \in \C$.
%Similarly to geometry of molecule, the geometry of ligand is union of hard sphere model.
%A probe is represented by one sphere of radius $\probe$.
%The collision-free region $\CF \subseteq \C$ is formed by configurations, where $\Sprobe$ can be placed without any collision, i.e., 
%$\CF = \{q \in \C | \Sprobe(q) \cap \SS = \emptyset\}$, where $\Sprobe(q)$ is the sphere with radius $\probe$ at configuration $q$.
%The distance $\dist(q_1,q_2)$  between two configurations $q_1,q_2\in\C$ is measured using 3D Euclidean metric.

%This radius can be detected using collision detection method.
%The configurations outside the protein $\CG \subseteq \C$ are collision-free with sphere $\Sgprobe$ of radius $\gprobe$, $\gprobe> \probe$,
%$\CG=\{q\in \C| \Sgprobe(q) \cap \SS = \emptyset \}$.
%A configuration $q \in \CG$ is referred to as {\sl goal configuration} in the rest of the paper.
%By computing $\alpha$-shape of the molecule, we can identify atoms located at the surface. 
%These atoms are referred to as {\sl surface atoms} in the rest of the paper.
%The distance between a configuration $q \in \C$ and its nearest surface atom is referred to as $\dists(q)$.
%We assume, that for each configuration $q$, the largest collision-free radius $r(q)$ can be computed.
%Let $r(q)$ denote the radius of the largest collision-free sphere at configuration $q$.

%Dynamic proteins are represented by a sequence of $n$ snapshots (frames).
%The tunnels change with the protein dynamics as well: their width and position change, they can merge with other tunnels or even disappear.
%The free region $\CF$ is therefore different in each frame, as positions of atoms change.
%For the sake of simplicity, a single notation $\C$ resp. $\CF$ is used in this paper and it is valid for the frame being processed.

%Approximated WVD diagram of the atoms is computed by representing each atom using 12 balls of equal radii and computing ordinary Voronoi diagram.
%Computing WVD this way is more efficient and numerically stable and it is used also in other related tools~\cite{caver3,yaffe2008}.
%Vertices of this WVD are filtered out if their distance to an atom is less than probe radius $\probe$.
%Let $\VV$ contain the remaining Voronoi vertices, and let $r_v \in \VV$ denote the distance to the nearest atom.
%The remaining Voronoi vertices $\VV$ of the WVD and their distances $r_v, v\in \VV$ to the nearest atom are then used in the following method.

@@

\linesnumbered
\begin{algorithm}[h]
%\setstretch{\straa}
\caption{\label{alg::tunnel}traversability planning}
\KwIn{
    TODO
}
\KwData{
    todo
}
\KwOut{
    todo
}
\hrule
$v = 0$; // index of the virtual goal\\
$iteration = 0$\;
\While{$iteration < \Imax$ and $v < N$}{
    \eIf{$rand() < \gb$}{
        $\qrand$ = random sample around $P[v]$\;
    }{
        $\qrand$ = random sample from $\C$\;
    }
    $\qnear$ = nearest node in $\T$ towards $\qrand$\;

    $S = $expandConfigurations($\qnear$)\;

    \For{$q \in S$}{
        $\T$.addNode($q$)\;
        $\T$.addEdge($\qnear, q$)\;
    }

    \For{$i=N : v$}{
        $d$ = nearest node in the tree to $P[i]$\;
        \If{$d < DT$}{
            $v = i$; // new virtual goal found\;
            break\;
        }
    }
}
trajectory = extract from tree\;
\end{algorithm}
